Operational Feature Selection with Gaussian Mixture Models for remote sensing images

### 2016-10-06 paper.tex
Fin section IV
+ passage en double colonne

\documentclass[journal,peerreview,onecolumn]{IEEEtran}
% \documentclass[journal]{IEEEtran}
\usepackage{etex}
\usepackage{cite}
\usepackage[pdftex]{graphicx}
\graphicspath{{./Fig/}}
\DeclareGraphicsExtensions{.pdf,.jpg,.png}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{array}
\usepackage[caption=false,font=footnotesize]{subfig}
\usepackage{fixltx2e}
\usepackage{url}
\usepackage{listings}
\usepackage{multirow}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage{tabularx}
\usepackage{empheq}
\usepackage{hyperref}
\usepackage{booktabs}

\newcommand\MyBox[2]{
  \fbox{\lower0.75cm
    \vbox to 1.2cm{\vfil
      \hbox to 1.2cm{\hfil\parbox{1.4cm}{#1\\#2}\hfil}
      \vfil}%
  }%
}

% for peerreview
\usepackage{setspace}
\onehalfspacing

@@

\tableofcontents
\clearpage
% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
This article presents a forward feature selection algorithm based on Gaussian mixture model (GMM) classifiers. The algorithm selects iteratively features that maximize a criterion function which can be either a classification rate or a measure of divergence. Several variations of this algorithm are explored by changing the criterion function and also by testing a floating forward variation allowing backward step to discard already selected features.

@@

Finally, the method is compared to standard classifiers using two different datasets, one of hyperspectral images with a lot of spectral variables and one with heterogeneous spatial features. The results validate the fact that the method performs well in terms of processing time and classification accuracy in comparison to the standard classifiers available in OTB.\\
\end{abstract}

@@

Features selection/extraction algorithms can be divided into three types. The first type, called \emph{filter methods}, is based uniquely on data analysis in the sense that the method is independant of any classifier. Features are ranked according to a statistical analysis of the data. For example, the Principal Component Analysis (PCA) described in \cite{jimenez1998supervised} is a typical unsupervised filter method. Bruzzone \emph{et al.}\cite{bruzzone1995extension} develop a supervised filter method based on Jeffries-Matusita distance to maximize the separability of class distribution. A filter based on the Pearson $\chi^2$ statistical test has been used on biomedical data\cite{biesiada2007feature}. The use of correlation between bands has been explored with hyperspectral data\cite{demir2008phase}. In general, these methods have the advantage to be fast and usable with any classifier but they do not take into account the strength and weakness of the chosen classifier and do not optimize directly the classification process.

@@

The remaining of this article is organized as follows. Section~\ref{sec:gmm-hd} presents Gaussian Mixture Model classifiers and the problems link to high-dimensionality. The features selection methods used to make them suitable for high-dimension space is detailed in Section~\ref{sec:selection}. The work done to develop an efficient implementation of the proposed selection method is presented in Section~\ref{sec:implementation}. And finally, after introducing the two datasets in Section~\ref{sec:datasets}, the tests conducted to explore variations of the algorithm and to compare it to other standard classifiers are shown and commented in Section~\ref{sec:test}.

@@

    \begin{figure}[!t]
        \centering
        \begin{tikzpicture}
          \begin{semilogyaxis}[xmin=0,xmax=200,ymin=0,width=0.5*\columnwidth,grid,axis x line=left ,axis y line=left, tick align=outside,xlabel = $d$,ylabel = $\eta_c$]
            \addplot+[very thick,mark=none,smooth,domain=0:200,samples=200] (\x,{\x*(\x+3)/2+1});
            \end{semilogyaxis}
        \end{tikzpicture}
        \caption{Number of parameters $\eta_c$ by class in function of dimension $d$: $\eta_c=d(d+3)/2+1$.\label{fig:nb-param}}
    \end{figure}

@@

        \emph{The   Kullback-Leibler   divergence}   (KL   divergence)
        measures  the  amount  of  information  lost  when  the  first
        distribution     is     approximated     by     the     second
        one\cite{kullback1987letter}. It can be explicitly computed in
        the case of Gaussian distributions:
        \begin{align}
            &KL_{ij} = \frac{1}{2} \biggl( \text{Tr} (\boldsymbol{\Sigma}_i^{-1} \boldsymbol{\Sigma}_j) \nonumber \\
            & + (\boldsymbol{\mu}_i - \boldsymbol{\mu}_j)^t \boldsymbol{\Sigma}_i^{-1} (\boldsymbol{\mu}_i - \boldsymbol{\mu}_j) - d + \log \left( \frac{|\boldsymbol{\Sigma}_i|}{|\boldsymbol{\Sigma}_j|} \right) \biggr),
        \end{align}
        where Tr is the trace operator and $d$ the dimension of the distribution.

        The    KL   divergence    is   not    symmetric,   \emph{i.e.}
        $KL_{ij} \ne KL_{ji}$.  A symmetrized version  is used to
        compute the criterion function:
        \begin{align}\label{eq:skl}
          SKL_{ij} &=KL_{ij} + KL_{i,j} \nonumber \\
            &= \frac{1}{2} \biggl( \text{Tr} (\boldsymbol{\Sigma}_i^{-1} \boldsymbol{\Sigma}_j + \boldsymbol{\Sigma}_j^{-1} \boldsymbol{\Sigma}_i) \nonumber \\
            &~~+ (\boldsymbol{\mu}_i - \boldsymbol{\mu}_j)^t (\boldsymbol{\Sigma}_i^{-1} + \boldsymbol{\Sigma}_j^{-1}) (\boldsymbol{\mu}_i - \boldsymbol{\mu}_j) - 2d \biggr).
        \end{align}

@@

        The \emph{Jeffries–Matusita distance} is a measure based on the Bhattacharyya distance. It saturates if the separability between the two distribution increases~\cite{bruzzone2009novel}. The JM distance is defined as
        \begin{equation}\label{eq:jm}
            {JM}_{ij} = \sqrt{ 2 \{1 - \text{exp}[-B_{ij}]\} }.
        \end{equation}

@@

        \begin{table}[!t]
            \centering
            \caption{Summary of the different criterion functions.\label{tab:crit}}
            \begin{tabular}[b]{lcc}
              \toprule
              Criterion & Type & Complexity \\
              \midrule
              Overall accuracy            & Accuracy   & High \\
              Cohen's kappa               & Accuracy   & High\\
              F1 mean                     & Accuracy   & High\\
              \midrule
              Kullback-Leibler divergence & Divergence & Low \\
              Jeffries-Matusita distance  & Divergence & Low \\
              \bottomrule
            \end{tabular}
        \end{table}

@@

\subsection{Statistical update rules}
        \subsubsection{Update for cross validation}
        \label{sec:update-cv}
        Based on \cite{fauvel2015fast}, a method to accelerate the $n_{cv}$-fold cross-validation process in the case of criterion functions based on correct classification measures was implemented. The idea is to estimate the GMM with the whole training set once and then, instead of training models on $(n_{cv}-1)$ folds, parameters of the complete model are used to derive those of submodels.

        A sub-GMM model can be estimated at a reduced complexity when one fold is remove from the training set. From~\cite{fauvel2015fast}, the following formulae for unbiased estimators can be derived: %% TODO: rajouter des \hat{} partout: ce sont des estimateurs.
        \begin{prop}
            \label{eq:update-cv1}
            (Cross-validation mean update)
            \begin{equation*}
                \boldsymbol{\mu}_c^{n_c-\nu_c} = \frac{n_c \boldsymbol{\mu}_c^{n_c} - \nu_c \boldsymbol{\mu}_c^{\nu_c}}{n_c - \nu_c} \nonumber
            \end{equation*}
        \end{prop}
        \begin{prop}
            \label{eq:update-cv2}
            (Cross-validation covariance matrix update)
            \begin{align*}
                \boldsymbol{\Sigma}_c^{n_c-\nu_c} = &\frac{1}{n_c-\nu_c-1} \biggl( (n_c-1) \boldsymbol{\Sigma}_c^{n_c} - (\nu_c-1) \boldsymbol{\Sigma}_c^{\nu_c} \nonumber \\
                &- \frac{n_c \nu_c}{(n_c-\nu_c)} (\boldsymbol{\mu}_c^{\nu_c}-\boldsymbol{\mu}_c^{n_c})(\boldsymbol{\mu}_c^{\nu_c}-\boldsymbol{\mu}_c^{n_c})^t \biggr) \nonumber
            \end{align*}
        \end{prop}
        \noindent where $n_c$ is the number of samples of class $c$, $\nu_c$ is the number of samples of class $c$ removed from the initial set, exposants on $\boldsymbol{\Sigma}_c$ and $\boldsymbol{\nu}_c$ denotes the set of samples used to compute them.

        \subsubsection{Criterion function computation}
        \label{sec:update-crit}
        At iteration $k$, depending on the criterion, three terms have
        to  be computed:  the inverse  of the  covariance matrix,  the
        logarithm of the determinant of  the covariance matrix and the
        quadratical term in  eq.~\ref{eq:decision}. However, all these
        terms  have already  been computed  for iteration  $(k-1)$. By
        using  the  fact that  the  covariance  matrix is  a  positive
        definite     symmetric    matrix     and    block     matrices
        formulae~\cite[Chapter   9.2]{webb2003statistical},    it   is
        possible to factorize these terms at iteration $k$.

@@

        \begin{equation}\label{eq:cov}
            \boldsymbol{\Sigma}^{(k)} =
            \bigg[\begin{array}{cc}
            \boldsymbol{\Sigma}^{(k-1)} & \mathbf{u}      \\
            \mathbf{u}^t          & \sigma_{kk} \\
            \end{array}\bigg],
        \end{equation}
        where  $\mathbf{u}$  is  the  $k^{th}$ column  of  the  matrix
        without      the      diagonal      element,      \emph{i.e.},
        $\mathbf{u}_{i}   =    \boldsymbol{\Sigma}^{(k)}_{i,k}$   with
        $i  \in [1,k-1]$.  Using  block matrix  inverse formulae,  the
        following proposition holds:
        \begin{prop}
        \label{eq:update-inv}
            (Forward update rule for inverse of covariance matrix)
            \begin{equation}\label{eq:cov:inv}
                (\boldsymbol{\Sigma}^{(k)})^{-1} =
                \bigg[\begin{array}{cc}
                \mathbf{A} & \mathbf{v} \\
                \mathbf{v}^t  & \frac{1}{\alpha} \\
                \end{array}\bigg]
            \end{equation}
        \end{prop}
        \noindent                                                where
        $\mathbf{A}     =     (\boldsymbol{\Sigma}^{(k-1)})^{-1}     +
        \frac{1}{\alpha} (\boldsymbol{\Sigma}^{(k-1)})^{-1} \mathbf{u}
        \mathbf{u}^t              (\boldsymbol{\Sigma}^{(k-1)})^{-1}$,
        $\mathbf{v}           =           -           \frac{1}{\alpha}
        (\boldsymbol{\Sigma}^{(k-1)})^{-1}       \mathbf{u}$       and
        $      \alpha      =      \sigma_{kk}      -      \mathbf{u}^t
        (\boldsymbol{\Sigma}^{(k-1)})^{-1}  \mathbf{u}$.  This  update
        formulae      is     used      to      obtain     all      the
        $(\boldsymbol{\Sigma}^{(k)})^{-1}$  corresponding  to all  the
        possible augmented set of a given selection iteration. Similar
        update formulae can be written for the backward step.

@@

        Using eq.~\ref{eq:cov} and~\ref{eq:cov:inv}, it is possible to
        deduce  the   following  propositions  (proof  are   given  in
        Appendix~\ref{app:proof-update}).
        \begin{prop}
        \label{eq:update-quad}
            (Update rule for quadratical term)
            \begin{align}
                (\mathbf{x}^{(k)})^t (\boldsymbol{\Sigma}^{(k)})^{-1} \mathbf{x}^{(k)} = &\underbrace{(\mathbf{x}^{(k-1)})^t (\boldsymbol{\Sigma}^{(k-1)})^{-1} \mathbf{x}^{(k-1)}}_{\substack{\text{computed once}\\ \text{per selection step}}} \nonumber \\
                &+ \underbrace{\alpha ( \left[\begin{array}{cc} \mathbf{v}^t & \frac{1}{\alpha} \end{array}\right] \mathbf{x}^{(k)} )^2}_{\substack{\text{computed for} \\ \text{each augmented set}}}.
            \end{align}
        \end{prop}
        \begin{prop}
        \label{eq:update-log}
            (Update rule for logdet)
            \begin{equation}
                \log \left(|\boldsymbol{\Sigma}^{(k)}|\right) = \underbrace{\log \left(|\boldsymbol{\Sigma}^{(k-1)}|\right)}_{\substack{\text{computed once}\\ \text{per selection step}}} + \underbrace{\log \alpha}_{\substack{\text{computed for} \\ \text{each augmented set}}}.
            \end{equation}
        \end{prop}

@@

      \begin{prop}[Kullback-Leibler divergence~\ref{eq:skl}]
        \begin{align}
        \label{eq:skl-update}
            & SKL_{ij} = \frac{1}{2} \biggl( \underbrace{\text{Tr} \left((\boldsymbol{\Sigma}_i^{(k)})^{-1} \boldsymbol{\Sigma}_j^{(k)} + (\boldsymbol{\Sigma}_j^{(k)})^{-1} \boldsymbol{\Sigma}_i^{(k)}\right)}_{\substack{\text{computed for} \\ \text{each augmented set}}} \nonumber \\
            &+ \underbrace{\alpha ( \left[\begin{array}{cc} \mathbf{v_i}^t & \frac{1}{\alpha_i} \end{array}\right] (\boldsymbol{\mu}_i^{(k)} - \boldsymbol{\mu}_j^{(k)}) )^2}_{\substack{\text{computed for} \\ \text{each augmented set}}} \nonumber \\
            &+ \underbrace{\alpha ( \left[\begin{array}{cc} \mathbf{v_j}^t & \frac{1}{\alpha_j} \end{array}\right] (\boldsymbol{\mu}_i^{(k)} - \boldsymbol{\mu}_j^{(k)}) )^2 - 2k}_{\substack{\text{computed for} \\ \text{each augmented set}}} \nonumber \\
            &+ \underbrace{(\boldsymbol{\mu}_i^{(k-1)} - \boldsymbol{\mu}_j^{(k-1)})^t ((\boldsymbol{\Sigma}_i^{(k-1)})^{-1} + (\boldsymbol{\Sigma}_j^{(k-1)})^{-1}) (\boldsymbol{\mu}_i^{(k-1)} - \boldsymbol{\mu}_j^{(k-1)})}_{\substack{\text{computed once}\\ \text{per selection step}}} \biggr).
        \end{align}
        with $(\boldsymbol{\Sigma}_i^{(k)})^{-1}$ computed with Proposition~\ref{eq:update-inv};
      \end{prop}

      \begin{prop}[Jeffries-Matusita distance~\ref{eq:jm}]
      \begin{align}
        \label{eq:jm-update}
            \text{B}_{ij} = &\underbrace{\frac{1}{4} (\boldsymbol{\mu}_i^{(k-1)} - \boldsymbol{\mu}_j^{(k-1)})^t ( \boldsymbol{\tilde{\Sigma}}^{(k-1)} )^{-1} (\boldsymbol{\mu}_i^{(k-1)} - \boldsymbol{\mu}_j^{(k-1)})}_{\substack{\text{computed once}\\ \text{per selection step}}} \nonumber \\
            &+ \underbrace{\frac{1}{2} \ln \left( \frac{|\boldsymbol{\tilde{\Sigma}}^{(k-1)}|}{\sqrt{|\boldsymbol{\Sigma}_i^{(k-1)}| |\boldsymbol{\Sigma}_j^{(k-1)}|}} \right)}_{\substack{\text{computed once}\\ \text{per selection step}}} \nonumber \\
            &+ \underbrace{\frac{1}{4} \tilde{\alpha} ( \left[\begin{array}{cc} \mathbf{\tilde{v}}^t & \frac{1}{\tilde{\alpha}} \end{array}\right] (\boldsymbol{\mu}_{i}^{(k)} - \boldsymbol{\mu}_{j}^{(k)}) )^2 + \frac{1}{2} \ln \left( \frac{\tilde{\alpha}}{\sqrt{\alpha_i \alpha_j}} \right)}_{\substack{\text{computed for} \\ \text{each augmented set}}},
      \end{align}
      where $\boldsymbol{\tilde{\Sigma}} = \boldsymbol{\Sigma}_i + \boldsymbol{\Sigma}_j$.
    \end{prop}

@@

    \subsection{OTB external module}
    \label{sec:otb-module}

    The developed method has been made available as a remote module of Orfeo toolbox called \emph{otbFFSforGMM} and can be download on Github\footnote{\url{https://github.com/Laadr/otbFFSforGMM}}.

    The Orfeo Toolbox is an open-source library for remote sensing image processing. This library allows developers to code external modules and make them available to the community. It is to note that the toolbox already implements several classifiers inherited mostly from OpenCV. Nevertheless, the GMM classifier currently in place is not robust to high dimensionality and that is why a new GMM model has been developed in the module.

    In addition of the C++ class directly available to develop C++ code, three commandline applications have been produced. These applications allow the user to use and parametrize the algorithm as desired.

    The first application \emph{otbcli\_TrainGMMApp} trains a GMM classifier from multiple pairs of images and training vector data and creates a model file. It also allows to perform ridge regularization with an embedded selection of the regularization parameter among a list of values proposed by the user. The regularization value is chosen by maximizing a criterion function (Cohen's kappa, overall accuracy or mean of F1 scores) estimated by cross-validation.

    The second application \emph{otbcli\_TrainGMMSelectionApp} also trains a GMM classifier and in addition performs a feature selection algorithm (SFS, SFFS) with various possible criterion (Jeffries-Matusita dist., Kullback-Lieber div., Cohen’s kappa, overall accuracy, mean F1-score).

>>>>

%\documentclass[journal,peerreview,onecolumn]{IEEEtran}
\documentclass[journal]{IEEEtran}
\usepackage{etex}
\usepackage{cite}
\usepackage[pdftex]{graphicx}
\graphicspath{{./Fig/}}
\DeclareGraphicsExtensions{.pdf,.jpg,.png}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{array}
\usepackage[caption=false,font=footnotesize]{subfig}
\usepackage{fixltx2e}
\usepackage{url}
\usepackage{listings}
\usepackage{multirow}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage{tabularx}
\usepackage{empheq}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{minted}\definecolor{bg}{rgb}{0.95,0.95,0.95}
\newcommand\MyBox[2]{
  \fbox{\lower0.75cm
    \vbox to 1.2cm{\vfil
      \hbox to 1.2cm{\hfil\parbox{1.4cm}{#1\\#2}\hfil}
      \vfil}%
  }%
}

% for peerreview
% \usepackage{setspace}
% \onehalfspacing

@@

% \tableofcontents
% \clearpage
% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
This article presents a forward feature selection algorithm based on Gaussian mixture model (GMM) classifiers. The algorithm selects iteratively features that maximize a criterion function which can be either a classification rate or a measure of divergence. Several variations of this algorithm are explored by changing the criterion function and also by testing a floating forward variation allowing backward step to discard already selected features.

@@

Finally, the method is compared to standard classifiers using two different data-sets, one of hyperspectral images with a lot of spectral variables and one with heterogeneous spatial features. The results validate the fact that the method performs well in terms of processing time and classification accuracy in comparison to the standard classifiers available in OTB.\\
\end{abstract}

@@

Features selection/extraction algorithms can be divided into three types. The first type, called \emph{filter methods}, is based uniquely on data analysis in the sense that the method is independent of any classifier. Features are ranked according to a statistical analysis of the data. For example, the Principal Component Analysis (PCA) described in \cite{jimenez1998supervised} is a typical unsupervised filter method. Bruzzone \emph{et al.}\cite{bruzzone1995extension} develop a supervised filter method based on Jeffries-Matusita distance to maximize the separability of class distribution. A filter based on the Pearson $\chi^2$ statistical test has been used on bio-medical data\cite{biesiada2007feature}. The use of correlation between bands has been explored with hyperspectral data\cite{demir2008phase}. In general, these methods have the advantage to be fast and usable with any classifier but they do not take into account the strength and weakness of the chosen classifier and do not optimize directly the classification process.

@@

The remaining of this article is organized as follows. Section~\ref{sec:gmm-hd} presents Gaussian Mixture Model classifiers and the problems link to high-dimensionality. The features selection methods used to make them suitable for high-dimension space is detailed in Section~\ref{sec:selection}. The work done to develop an efficient implementation of the proposed selection method is presented in Section~\ref{sec:implementation}. And finally, after introducing the two data-sets in Section~\ref{sec:datasets}, the tests conducted to explore variations of the algorithm and to compare it to other standard classifiers are shown and commented in Section~\ref{sec:test}.

@@

    \begin{figure}[!t]
        \centering
        \begin{tikzpicture}
          \begin{axis}[ymode=log,xmin=0,xmax=200,ymin=0,width=0.8\columnwidth,grid,axis x line=left ,axis y line=left, tick align=outside,xlabel = $d$,ylabel = $\eta_c$,small]
            \addplot+[very thick,mark=none,smooth,domain=0:200,samples=200] (\x,{\x*(\x+3)/2+1});
            \end{axis}
        \end{tikzpicture}
        \caption{Number of parameters $\eta_c$ by class in function of dimension $d$: $\eta_c=d(d+3)/2+1$.\label{fig:nb-param}}
    \end{figure}

@@

        \emph{The   Kullback-Leibler   divergence}   (KL   divergence)
        measures  the  amount  of  information  lost  when  the  first
        distribution     is     approximated     by     the     second
        one\cite{kullback1987letter}. It can be explicitly computed in
        the case of Gaussian distributions:
        \begin{align}
            &KL_{ij} = \frac{1}{2} \biggl\{ \text{Tr} (\boldsymbol{\Sigma}_i^{-1} \boldsymbol{\Sigma}_j) \nonumber \\
            & + (\boldsymbol{\mu}_i - \boldsymbol{\mu}_j)^t \boldsymbol{\Sigma}_i^{-1} (\boldsymbol{\mu}_i - \boldsymbol{\mu}_j) - d + \log \left( \frac{|\boldsymbol{\Sigma}_i|}{|\boldsymbol{\Sigma}_j|} \right) \biggr\},
        \end{align}
        where Tr is the trace operator and $d$ the dimension of the distribution.

        The    KL   divergence    is   not    symmetric,   \emph{i.e.}
        $KL_{ij} \ne KL_{ji}$.  A symmetrized version  is used to
        compute the criterion function:
        \begin{align}\label{eq:skl}
          SKL_{ij} &=KL_{ij} + KL_{i,j} \nonumber \\
            &= \frac{1}{2} \biggl\{ \text{Tr} (\boldsymbol{\Sigma}_i^{-1} \boldsymbol{\Sigma}_j + \boldsymbol{\Sigma}_j^{-1} \boldsymbol{\Sigma}_i) \nonumber \\
            &~~+ (\boldsymbol{\mu}_i - \boldsymbol{\mu}_j)^t (\boldsymbol{\Sigma}_i^{-1} + \boldsymbol{\Sigma}_j^{-1}) (\boldsymbol{\mu}_i - \boldsymbol{\mu}_j) - 2d \biggr\}.
        \end{align}

@@

        The \emph{Jeffries–Matusita distance} is a measure based on the Bhattacharyya distance. It saturates if the separability between the two distribution increases~\cite{bruzzone2009novel}. The JM distance is defined as
        \begin{equation}\label{eq:jm}
            {JM}_{ij} = \sqrt{ 2 \{1 - \text{exp}(-B_{ij})\} }.
        \end{equation}

@@

        \begin{table}[!t]
            \centering
            \caption{Summary of the different criterion functions.\label{tab:crit}}
            \begin{tabular}[b]{lcc}
              \toprule
              Criterion & Type & Complexity \\
              \midrule
              Overall accuracy            & Accuracy   & High \\
              Cohen's kappa               & Accuracy   & High\\
              F1 mean                     & Accuracy   & High\\
              \midrule
              Kullback-Leibler divergences & Divergence & Low \\
              Jeffries-Matusita distance  & Divergence & Low \\
              \bottomrule
            \end{tabular}
        \end{table}

@@

\subsection{Statistical update rules}
        \subsubsection{Update for cross validation}
        \label{sec:update-cv}
        Based on \cite{fauvel2015fast}, a method to accelerate the $n_{cv}$-fold cross-validation process in the case of criterion functions based on correct classification measures was implemented. The idea is to estimate the GMM with the whole training set once and then, instead of training models on $(n_{cv}-1)$ folds, parameters of the complete model are used to derive those of sub-models.

        A sub-GMM model can be estimated at a reduced complexity when one fold is remove from the training set. From~\cite{fauvel2015fast}, the following formulae for unbiased estimators can be derived: %% TODO: rajouter des \hat{} partout: ce sont des estimateurs.
        \begin{prop}
            \label{eq:update-cv1}
            (Cross-validation mean update)
            \begin{equation*}
                \boldsymbol{\mu}_c^{n_c-\nu_c} = \frac{n_c \boldsymbol{\mu}_c^{n_c} - \nu_c \boldsymbol{\mu}_c^{\nu_c}}{n_c - \nu_c} \nonumber
            \end{equation*}
        \end{prop}
        \begin{prop}
            \label{eq:update-cv2}
            (Cross-validation covariance matrix update)
            \begin{align*}
                \boldsymbol{\Sigma}_c^{n_c-\nu_c} = &\frac{1}{n_c-\nu_c-1} \biggl\{ (n_c-1) \boldsymbol{\Sigma}_c^{n_c} - (\nu_c-1) \boldsymbol{\Sigma}_c^{\nu_c} \nonumber \\
                &- \frac{n_c \nu_c}{(n_c-\nu_c)} (\boldsymbol{\mu}_c^{\nu_c}-\boldsymbol{\mu}_c^{n_c})(\boldsymbol{\mu}_c^{\nu_c}-\boldsymbol{\mu}_c^{n_c})^t \biggr\} \nonumber
            \end{align*}
        \end{prop}
        \noindent where $n_c$ is the number of samples of class $c$, $\nu_c$ is the number of samples of class $c$ removed from the initial set, exponents on $\boldsymbol{\Sigma}_c$ and $\boldsymbol{\nu}_c$ denotes the set of samples used to compute them.

        \subsubsection{Criterion function computation}
        \label{sec:update-crit}
        At iteration $k$, depending on the criterion, three terms have
        to  be computed:  the inverse  of the  covariance matrix,  the
        logarithm of the determinant of  the covariance matrix and the
        quadratic  term in  eq.~\ref{eq:decision}. However,  all these
        terms  have already  been computed  for iteration  $(k-1)$. By
        using  the  fact that  the  covariance  matrix is  a  positive
        definite     symmetric    matrix     and    block     matrices
        formulae~\cite[Chapter   9.2]{webb2003statistical},    it   is
        possible to factorize these terms at iteration $k$.

@@

        \begin{equation}\label{eq:cov}
            \boldsymbol{\Sigma}^{(k)} =
            \bigg[\begin{array}{cc}
            \boldsymbol{\Sigma}^{(k-1)} & \mathbf{u}      \\
            \mathbf{u}^t          & \sigma_{kk} \\
            \end{array}\bigg],
        \end{equation}
        where  $\mathbf{u}$  is  the  $k^{th}$ column  of  the  matrix
        without      the      diagonal      element,      \emph{i.e.},
        $\mathbf{u}_{i}   =    \boldsymbol{\Sigma}^{(k)}_{i,k}$   with
        $i  \in [1,k-1]$.  Using  block matrix  inverse formulae,  the
        following proposition holds:
        \begin{prop}
        \label{eq:update-inv}
            (Forward update rule for the inverse of the covariance matrix)
            \begin{equation}\label{eq:cov:inv}
                (\boldsymbol{\Sigma}^{(k)})^{-1} =
                \bigg[\begin{array}{cc}
                \mathbf{A} & \mathbf{v} \\
                \mathbf{v}^t  & \frac{1}{\alpha} \\
                \end{array}\bigg]
            \end{equation}
        \end{prop}
        \noindent                                                where
        $\mathbf{A}     =     (\boldsymbol{\Sigma}^{(k-1)})^{-1}     +
        \frac{1}{\alpha} (\boldsymbol{\Sigma}^{(k-1)})^{-1} \mathbf{u}
        \mathbf{u}^t              (\boldsymbol{\Sigma}^{(k-1)})^{-1}$,
        $\mathbf{v}           =           -           \frac{1}{\alpha}
        (\boldsymbol{\Sigma}^{(k-1)})^{-1}       \mathbf{u}$       and
        $      \alpha      =      \sigma_{kk}      -      \mathbf{u}^t
        (\boldsymbol{\Sigma}^{(k-1)})^{-1}  \mathbf{u}$.  This  update
        formulae      is     used      to      obtain     all      the
        $(\boldsymbol{\Sigma}^{(k)})^{-1}$  corresponding  to all  the
        possible augmented set of a given selection iteration. Similar
        update formulae can be written for the backward step.

@@

        Using eq.~\ref{eq:cov} and~\ref{eq:cov:inv}, it is possible to
        deduce  the   following  propositions  (proof  are   given  in
        Appendix~\ref{app:proof-update}).
        \begin{prop}
        \label{eq:update-quad}
            (Update rule for the quadratic term)
            \begin{align}
              (\mathbf{x}^{(k)})^t (\boldsymbol{\Sigma}^{(k)})^{-1} \mathbf{x}^{(k)} = &\underbrace{(\mathbf{x}^{(k-1)})^t (\boldsymbol{\Sigma}^{(k-1)})^{-1} \mathbf{x}^{(k-1)}}_{\substack{\text{computed once per selection step}}} \nonumber \\
                                                                                       &+ \underbrace{\alpha ( \left[\begin{array}{cc} \mathbf{v}^t & \frac{1}{\alpha} \end{array}\right] \mathbf{x}^{(k)} )^2}_{\substack{\text{computed for each augmented set}}}.
            \end{align}
        \end{prop}
        \begin{prop}
        \label{eq:update-log}
            (Update rule for logdet)
            \begin{equation}
                \log \left(|\boldsymbol{\Sigma}^{(k)}|\right) = \underbrace{\log \left(|\boldsymbol{\Sigma}^{(k-1)}|\right)}_{\substack{\text{computed once}\\ \text{per selection step}}} + \underbrace{\log \alpha}_{\substack{\text{computed for} \\ \text{each augmented set}}}.
            \end{equation}
        \end{prop}

@@

      \begin{prop}[Kullback-Leibler divergence~\ref{eq:skl}]
        \begin{align}
        \label{eq:skl-update}
            & SKL_{ij} = \frac{1}{2} \biggl\{ \underbrace{\text{\normalfont Tr} \left((\boldsymbol{\Sigma}_i^{(k)})^{-1} \boldsymbol{\Sigma}_j^{(k)} + (\boldsymbol{\Sigma}_j^{(k)})^{-1} \boldsymbol{\Sigma}_i^{(k)}\right)}_{\substack{\text{computed for each augmented set}}} \nonumber \\
            &+ \underbrace{\alpha ( \left[\begin{array}{cc} \mathbf{v_i}^t & \frac{1}{\alpha_i} \end{array}\right] (\boldsymbol{\mu}_i^{(k)} - \boldsymbol{\mu}_j^{(k)}) )^2}_{\substack{\text{computed for each augmented set}}} \nonumber \\
            &+ \underbrace{\alpha ( \left[\begin{array}{cc} \mathbf{v_j}^t & \frac{1}{\alpha_j} \end{array}\right] (\boldsymbol{\mu}_i^{(k)} - \boldsymbol{\mu}_j^{(k)}) )^2 - 2k}_{\substack{\text{computed for each augmented set}}} \nonumber \\
            &+ \underbrace{(\boldsymbol{\mu}_i^{(k-1)} - \boldsymbol{\mu}_j^{(k-1)})^t (\boldsymbol{\Sigma}_i^{(k-1)})^{-1}(\boldsymbol{\mu}_i^{(k-1)} - \boldsymbol{\mu}_j^{(k-1)})}_{\substack{\text{computed once per selection step}}} \nonumber\\
          & + \underbrace{(\boldsymbol{\mu}_i^{(k-1)} - \boldsymbol{\mu}_j^{(k-1)})^t(\boldsymbol{\Sigma}_j^{(k-1)})^{-1} (\boldsymbol{\mu}_i^{(k-1)} - \boldsymbol{\mu}_j^{(k-1)})}_{\substack{\text{computed once per selection step}}} \biggr\}.
        \end{align}
        with $(\boldsymbol{\Sigma}_i^{(k)})^{-1}$ computed with Proposition~\ref{eq:update-inv};
      \end{prop}

      \begin{prop}[Jeffries-Matusita distance~\ref{eq:jm}]
      \begin{align}
        \label{eq:jm-update}
            \text{B}_{ij} = &\underbrace{\frac{1}{4} (\boldsymbol{\mu}_i^{(k-1)} - \boldsymbol{\mu}_j^{(k-1)})^t ( \boldsymbol{\tilde{\Sigma}}^{(k-1)} )^{-1} (\boldsymbol{\mu}_i^{(k-1)} - \boldsymbol{\mu}_j^{(k-1)})}_{\substack{\text{computed once per selection step}}} \nonumber \\
            &+ \underbrace{\frac{1}{2} \ln \left( \frac{|\boldsymbol{\tilde{\Sigma}}^{(k-1)}|}{\sqrt{|\boldsymbol{\Sigma}_i^{(k-1)}| |\boldsymbol{\Sigma}_j^{(k-1)}|}} \right)}_{\substack{\text{computed once per selection step}}} \nonumber \\
            &+ \underbrace{\frac{1}{4} \tilde{\alpha} ( \left[\begin{array}{cc} \mathbf{\tilde{v}}^t & \frac{1}{\tilde{\alpha}} \end{array}\right] (\boldsymbol{\mu}_{i}^{(k)} - \boldsymbol{\mu}_{j}^{(k)}) )^2 + \frac{1}{2} \ln \left( \frac{\tilde{\alpha}}{\sqrt{\alpha_i \alpha_j}} \right)}_{\substack{\text{computed for each augmented set}}},
      \end{align}
      where $\boldsymbol{\tilde{\Sigma}} = \boldsymbol{\Sigma}_i + \boldsymbol{\Sigma}_j$.
    \end{prop}

@@

    \subsection{Implementation}
    \label{sec:otb-module}

    The proposed  method has been implemented  in \texttt{C++} through
    the  Orfeo  Toolbox (OTB)~\cite{christophe2008orfeo}.   The  Orfeo
    Toolbox  is  an  open-source  library  for  remote  sensing  image
    processing,  developed by  the  French Space  Agency (CNES).   The
    feature selection algorithm  can be installed as  a remote module,
    the           source            code           is           freely
    available\footnote{\url{https://github.com/Laadr/otbFFSforGMM}}.

    Following OTB framework, two application are available. The first,
    called  \texttt{otbcli\_TrainGMMSelectionApp}, performs  a feature
    selection algorithm (SFS or SFFS) with the one given criterion (to
    choose  between  Jeffries-Matusita  dist.,  Kullback-Lieber  div.,
    Cohen’s  kappa, overall  accuracy,  mean  F1-score).  The  second,
    called \texttt{otbcli\_PredictGMMApp}, generates the thematic maps
    according to the  learn model. The only  technical limitation that
    the  training set  must  fit  in the  RAM  of  the computer.   The
    classification step is streamed and there is no limitation in term
    of image  size. The Figure~\ref{fig:otb:ffs} shows  a code excerpt
    to run the application.

    \begin{figure}\centering

    \begin{minted}[fontsize=\tiny,bgcolor=bg]{sh}
otbcli_TrainGMMSelectionApp -io.il hyper.tif \
			    -io.vd reference.shp \
			    -gmm.varnb 20 -gmm.method forward -gmm.crit jm\
			    -gmm.best 1 -gmm.seed 0\
			    -io.out model.txt
